---
title: "covid-19-analysis"
author: "Anonymous Potato"
date: "2024-02-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libraries}

library(tidyverse)
library(lubridate)
```

# Introduction

This project was completed for DTSA5301 as part of the MSc-DS at the University of Colorado, Boulder. The data was acquired from the links below and contains the following information:

* time_series_covid19_confirmed_global.csv: Date, Province/State, Country/Region, number of confirmed cases.
* time_series_covid19_deaths_global.csv: Date, Province/State, Country/Region, number of deaths.

```{r import_data}
# reference 1

# base url 
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

# file urls
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv")

# population data lookup
uid <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

urls <- str_c(url_in,file_names)
```

```{r read_data, message=FALSE}

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
global_population <- read_csv(uid)
```

```{r tidy_global_cases}

global_cases <-global_cases %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', 
                         Lat, 
                         Long),
               names_to = 'date', 
               values_to = 'cases') %>% 
  select(-c(Lat, Long))

```

```{r tidy_global_deaths}

global_deaths <-global_deaths %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', 
                         Lat, 
                         Long),
               names_to = 'date', 
               values_to = 'deaths') %>% 
  select(-c(Lat, Long))
```


```{r concat_global_df, date_series}

global <- global_cases %>% 
  full_join(global_deaths) %>% 
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>% 
  mutate(date = mdy(date))
```

```{r}
summary(global)
```

```{r}

global <- global %>% 
  filter(cases > 0)

summary(global)
```

```{r}
canada <- global %>% 
  filter(Country_Region == 'Canada')

canada
```


```{r unique_prov}

unique(canada$Province_State)

```

```{r clean_prov}

canada <- subset(canada, !(Province_State %in% c('Diamond Princess', 'Grand Princess', 'Repatriated Travellers')))
```

```{r canada_pop}

canada_pop <- global_population %>% 
  filter(Country_Region == 'Canada') %>% 
  select(Province_State, Country_Region, Population)

canada_pop <- subset(canada_pop, !(Province_State %in% c('Diamond Princess', 'Grand Princess', 'Recovered')))

canada_pop
```

```{r}
canada <- canada %>% 
  left_join(canada_pop, by = 'Province_State') %>% 
  select(-c(Country_Region.x, Country_Region.y))

canada
```




```{r plot_natl_cases}

natl_cases <- canada %>% 
  group_by(date) %>% 
  summarise(total = sum(cases, na.rm = TRUE))


cases <- ggplot(natl_cases, aes(x = date, y = total)) +
  geom_point(colour = 'darkblue') +
  xlab('Date') +
  ylab('Number of cases (linear scale)') +
  ggtitle('Fig. 1: Cumulative Covid-19 cases in Canada, Feb 2020 - Mar 2023')

cases
```


```{r plot_prov_cases}

prov_cases <- ggplot(canada, aes(x = date, y = cases)) +
  geom_line(size = 1.5, colour = 'darkblue') +
  scale_y_log10() +
  facet_wrap(~ Province_State) +
  xlab('Date') +
  ylab('Number of cases (logarithmic scale)') +
  ggtitle('Fig. 2: Cumulative Covid-19 cases in Canada by province, Feb 2020 - Mar 2023')

prov_cases
```

```{r plot_natl_deaths}

# filter out 0 values so that log scale will work
canada_no_deaths <- canada %>% 
  filter(deaths >0)

# sum provincial daily deaths
natl_deaths <- canada_no_deaths %>% 
  group_by(date) %>% 
  summarise(total = sum(deaths, na.rm = TRUE))

# plot
deaths <- ggplot(natl_deaths, aes(x = date, y = total)) +
  geom_point(colour = 'darkred') +
  xlab('Date') +
  ylab('Number of deaths (linear scale)') +
  ggtitle('Fig. 3: Cumulative Covid-19 deaths in Canada, Feb 2020 - Mar 2023')

deaths
```

```{r }
prov_deaths <- ggplot(canada_no_deaths, aes(x = date, y = deaths)) +
  geom_line(size = 1.5, colour = 'darkred') +
  facet_wrap(~ Province_State) +
  scale_y_log10() +
  xlab('Date') +
  ylab('Number of deaths (logarithmic scale)') +
  ggtitle('Fig. 4: Covid-19 deaths in Canada by province, Feb 2020 - Mar 2023')

prov_deaths
```


```{r lm}

model_1 <- lm(deaths ~ cases, data = canada)

summary(model_1)
```

```{r predict}

canada$predictions <- predict(model_1)
```


```{r}
mod_1_plot <- ggplot(canada) +
  geom_point(aes(x = cases, y = deaths, colour = 'Actual'), alpha = 0.5) +
  geom_point(aes(x = cases, y = predictions, colour = 'Predicted'), size = 3) +
  labs(x = 'Cases', y = 'Deaths', 
       title = 'Fig. 5: Observed vs. predicted deaths per number of cases') +
  scale_colour_manual(values = c('Actual' = 'darkorange', 'Predicted' = 'blue'),
                      labels = c('Actual', 'Predicted')) +
       
  theme_minimal()  

mod_1_plot
```

```{r}
canada <- canada %>% 
  arrange(Province_State, date) %>% 
  group_by(Province_State) %>% 
  mutate(new_cases = cases - lag(cases, default = first(cases)))

canada
```

```{r}
canada <- canada %>% 
  arrange(Province_State, date) %>% 
  group_by(Province_State) %>% 
  mutate(new_deaths = deaths - lag(deaths, default = first(deaths)))

head(canada)
```
###### Removal of outliers

```{r}
canada <- subset(canada, !(new_cases < 0 | new_deaths < 0))

```

```{r}

canada <- subset(canada, !(Province_State == 'Ontario' & date == '2022-04-13'))
```


```{r}

model_2 <- lm(new_deaths ~ new_cases, data = canada)

summary(model_2)
```

```{r}
canada$new_predictions <- predict(model_2)

canada
```

```{r}

mod_2_plot <- ggplot(canada) +
  geom_point(aes(x = new_cases, y = new_deaths, colour = 'Actual'), alpha = 0.2) +
  geom_point(aes(x = new_cases, y = new_predictions, colour = 'Predicted'), size = 3) +
  labs(x = 'Cases', y = 'Deaths', 
       title = 'Fig. 5: Observed vs. predicted deaths per number of cases') +
  scale_colour_manual(values = c('Actual' = 'darkred', 'Predicted' = 'blue'),
                      labels = c('Actual', 'Predicted')) +
       
  theme_minimal()  

mod_2_plot
  
```

```{r}

library(plotly)
```

```{r}

new_cases_deaths <- ggplot(canada, aes(x = date)) +
  geom_bar(aes(y = new_cases), stat = 'identity', fill = 'blue') +
  geom_bar(aes(y = new_deaths), stat = 'identity', fill = 'darkred', alpha = 1) +
  labs(title = 'Fig. 6: New cases and deaths over time',
      x = 'Date',
      y = 'Count (logarithmic scale)') + 
  theme_minimal() +
  scale_y_continuous(trans = 'log10')

interactive <- ggplotly(new_cases_deaths)

interactive
```




# References

1. Code copied directly from dtsa5301 course lectures & materials

# Environment

```{r}
sessionInfo()
```










