---
title: "covid-19-analysis"
author: "Anonymous Potato"
date: "2024-02-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libraries}

library(tidyverse)
library(lubridate)
```

# Introduction

This project was completed for DTSA5301 as part of the MSc-DS at the University of Colorado, Boulder. The data was acquired from the links below and contains the following information:

* time_series_covid19_confirmed_global.csv: Date, Province/State, Country/Region, number of confirmed cases.
* time_series_covid19_deaths_global.csv: Date, Province/State, Country/Region, number of deaths.

```{r import_data}
# reference 1

# base url 
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

# file urls
file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv")

# population data lookup
uid <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

urls <- str_c(url_in,file_names)
```

```{r read_data, message=FALSE}

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
global_population <- read_csv(uid)
```

```{r tidy_global_cases}

global_cases <-global_cases %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', 
                         Lat, 
                         Long),
               names_to = 'date', 
               values_to = 'cases') %>% 
  select(-c(Lat, Long))

```

```{r tidy_global_deaths}

global_deaths <-global_deaths %>% 
  pivot_longer(cols = -c('Province/State', 
                         'Country/Region', 
                         Lat, 
                         Long),
               names_to = 'date', 
               values_to = 'deaths') %>% 
  select(-c(Lat, Long))
```


```{r concat_global_df, date_series}

global <- global_cases %>% 
  full_join(global_deaths) %>% 
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>% 
  mutate(date = mdy(date))
```

```{r}
summary(global)
```

```{r}

global <- global %>% 
  filter(cases > 0)

summary(global)
```

```{r}
canada <- global %>% 
  filter(Country_Region == 'Canada')

canada
```


```{r unique_prov}

unique(canada$Province_State)

```

```{r clean_prov}

canada <- subset(canada, !(Province_State %in% c('Diamond Princess', 'Grand Princess', 'Repatriated Travellers')))
```

```{r canada_pop}

canada_pop <- global_population %>% 
  filter(Country_Region == 'Canada') %>% 
  select(Province_State, Country_Region, Population)

canada_pop <- subset(canada_pop, !(Province_State %in% c('Diamond Princess', 'Grand Princess', 'Recovered')))

canada_pop
```

```{r}
canada <- canada %>% 
  left_join(canada_pop, by = 'Province_State') %>% 
  select(-c(Country_Region.x, Country_Region.y))

canada
```


```{r plot_natl_cases}

canada_no_deaths <- canada %>% 
  filter(deaths > 0)

natl_deaths <- canada %>% 
  group_by(date) %>% 
  summarise(total_deaths = sum(deaths, na.rm = TRUE))

natl_cases <- canada %>% 
  group_by(date) %>% 
  summarise(total_cases= sum(cases, na.rm = TRUE))

combined_data <- merge(natl_cases, natl_deaths, by = 'date', all = TRUE)

combined_plot <- ggplot(combined_data, aes(x = date)) +
  geom_point(aes(y = total_cases, colour = 'Cases'), size = 1.5) +
  geom_point(aes(y = total_deaths, colour = 'Deaths'), size = 1.5) +
  scale_y_log10() +
  scale_color_manual(values = c("Cases" = "darkblue", "Deaths" = "darkred")) +
  xlab('Date') +
  ylab('Number of cases (logarithmic scale)') +
  ggtitle('Fig. 1: Cumulative Covid-19 cases and deaths in Canada, Feb 2020 - Mar 2023') +
  theme(legend.title = element_blank())

combined_plot
```


```{r plot_prov_cases}

prov_cases <- ggplot(canada) +
  geom_line(aes(x = date, y = cases), size = 1.5, colour = 'darkblue') +
  geom_line(aes(x = date, y = deaths), size = 1.5, colour = 'darkred') +
  scale_y_log10() +
  facet_wrap(~ Province_State) +
  xlab('Date') +
  ylab('Number of cases (logarithmic scale)') +
  ggtitle('Fig. 2: Cumulative Covid-19 cases in Canada by province, Feb 2020 - Mar 2023')

prov_cases
```
# Modelling

Here, I'll model the number of covid-19 deaths as a function of cases.

```{r lm}

model <- lm(total_deaths ~ total_cases, data = combined_data)

summary(model)
```
Here the R-squared value is 0.9382, meaning about 94%% of the variance in the number of deaths is predictable from the number of number of given cases.

Now let's see how the predictions for this model fit the data set:

```{r predict}

combined_data$predictions <- predict(model)
```


```{r}
mod_plot <- ggplot(combined_data) +
  geom_point(aes(x = total_cases, y = total_deaths, colour = 'Actual'), alpha = 0.5) +
  geom_point(aes(x = total_cases, y = predictions, colour = 'Predicted'), size = 3) +
  labs(x = 'Cases', y = 'Deaths', 
       title = 'Fig. 5: Observed vs. predicted deaths per number of cases') +
  scale_colour_manual(values = c('Actual' = 'darkorange', 'Predicted' = 'purple'),
                      labels = c('Actual', 'Predicted')) +
       
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

mod_plot
```
In figure 5 we can see that the fit of the predicted cases is a fairly close to the actual values.

```{r}
#canada <- canada %>% 
#  arrange(Province_State, date) %>% 
#  group_by(Province_State) %>% 
#  mutate(new_cases = cases - lag(cases, default = first(cases)))

#canada
```

```{r}
#canada <- canada %>% 
#  arrange(Province_State, date) %>% 
#  group_by(Province_State) %>% 
#  mutate(new_deaths = deaths - lag(deaths, default = first(deaths)))

#head(canada)
```
###### Removal of outliers
Outliers were found through exploratory analysis

```{r}
canada <- subset(canada, !(new_cases < 0 | new_deaths < 0))

```

```{r}

canada <- subset(canada, !(Province_State == 'Ontario' & date == '2022-04-13'))
```

```{r}

library(plotly)
```

```{r}

new_cases_deaths <- ggplot(canada, aes(x = date)) +
  geom_bar(aes(y = new_cases, fill = 'New Cases'), stat = 'identity', show.legend = TRUE) +
  geom_bar(aes(y = new_deaths, fill = 'New Deaths'), stat = 'identity', show.legend = TRUE, alpha = 1) +
  labs(title = 'Fig. 6: Interactive plot - New cases and deaths over time',
      x = 'Date',
      y = 'Count (logarithmic scale)',
      fill = 'Legend') + 
  theme_minimal() +
  scale_y_continuous(trans = 'log10') +
  scale_fill_manual(values = c("blue", "darkred"), 
                    labels = c("New Cases", "New Deaths")) 

interactive <- ggplotly(new_cases_deaths)

print(interactive)
```

Figure 6 shows that new deaths follows a similiar pattern to the new cases, but with a lower 'amplitude'. It also appears that the 'peaks' in the deaths don't line up directly under the 'peaks' for new cases, but are shifted ever-so-slightly to the left . I interpret this as there is a time delay between getting the virus and a person dying from it. This raises another question - what would be the time lag between new cases and new deaths? This question, unfortunately, will have to be put off for further study.


# References

1. Code copied directly from dtsa5301 course lectures & materials

# Environment

```{r}
sessionInfo()
```










